---
title: "Mauna Loa CO2 Time Series Analysis"
author: "Brian Pham"
date: "2025-03-07"
output:
  pdf_document: default
  html_document: default
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=6, warning=FALSE, tidy=TRUE, message = FALSE, tidy.opts=list(width.cutoff=60))
```

```{r}
# adding libraries
library(tidyverse)
library(fpp3)
library(ggplot2)
library(dplyr)
library(nortest)
library(tseries)



file <- "co2_mm_gl.csv"
climate_data <- read_csv(file, skip = 38)
head(climate_data)

# filter data to only before 2020
climate_data_clean <- climate_data %>%
  filter(year < 2020)
```

```{r}
# plot of CO2 over the years
ggplot(climate_data_clean, aes(x = decimal, y = average)) +
geom_line(color = "blue") +
labs(
title = "Mauna Loa CO2 Concentration Over Time",
x = "Year",
y = "CO2 (ppm)"
) +
theme_minimal()

# dividing the data by month
climate_data_ts <- tsibble(month = yearmonth(seq(as.Date("1979-01-01"),
                                                 as.Date("2019-12-01"),
                                                 by = "1 month")),
                           Time = climate_data_clean$decimal,
                           CO2 = climate_data_clean$average,
                           index = month)


Month <- as_factor(month(climate_data_ts$month))

climate_data_ts <- climate_data_ts %>% mutate(LogCO2 = log(CO2))

# plot of Log(CO2) over the years by month

mean_LogCO2 <- mean(climate_data_ts$LogCO2)
climate_data_ts %>% autoplot(LogCO2) +
  geom_point(aes(y=LogCO2, color=Month)) +
  scale_color_manual(values = c("black", "blue", "purple", "red", "orange", "darkred", "green", "chartreuse4", 
                                "chocolate", "gray60", "gold4", "salmon")) +
  geom_hline(aes(yintercept = mean_LogCO2), lty=2) +
  ggtitle("Log(CO2) vs. Month") + xlab("Month") + ylab("Log(CO2)") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))
```
Visually the plot appears non-stationary but checking through the tests.

```{r}

time_series_components <- climate_data_ts %>% 
  model(STL(LogCO2 ~ season(window="periodic"))) %>%
  components()

climate_data_ts <- climate_data_ts %>% 
  add_column(seasonal = time_series_components$season_year,
             LogCO2_SA = time_series_components$season_adjust)

autoplot(time_series_components)


mean_LogCO2 <- mean(climate_data_ts$LogCO2_SA)
climate_data_ts %>% autoplot(LogCO2_SA) +
  geom_point(aes(y=LogCO2_SA, color=Month)) +
  scale_color_manual(values = c("black", "blue", "purple", "red", "orange", "darkred", "green", "chartreuse4", 
                                "chocolate", "gray60", "gold4", "salmon")) +
  geom_hline(aes(yintercept = mean_LogCO2), lty=2) +
  ggtitle("Log(CO2) vs. Month") + xlab("Month") + ylab("Log(CO2)") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

climate_data_ts %>% ACF(LogCO2_SA) %>% autoplot()
unitroot_kpss(climate_data_ts$LogCO2_SA)
adf.test(climate_data_ts$LogCO2_SA)
```
The time series is clearly non-stationary because the p-value from the KPSS test is 0.01 which is less than 0.05. (not sure how to interpret Kpss test, need to add things about the ACF plot) And the large p-value of ADF test is evidence that the time series is non-stationary.

```{r}
# Computing first differences
climate_data_ts <- climate_data_ts %>% mutate(diff_LogCO2_SA = difference(LogCO2_SA))
head(climate_data_ts)
tail(climate_data_ts)

mean_diff_LogCO2_SA <- mean(climate_data_ts$diff_LogCO2_SA)

climate_data_ts %>% autoplot(diff_LogCO2_SA) +
  geom_point(aes(y=diff_LogCO2_SA, color=Month)) +
  scale_color_manual(values = c("black", "blue", "purple", "red", "orange", "darkred", "green", "chartreuse4", 
                                "chocolate", "gray60", "gold4", "salmon")) +
  geom_hline(aes(yintercept = mean_diff_LogCO2_SA), lty=2) +
  ggtitle("Log(CO2) vs. Month") + xlab("Month") + ylab("Log(CO2)") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))
climate_data_ts %>% ACF(diff_LogCO2_SA) %>% autoplot()
unitroot_kpss(climate_data_ts$diff_LogCO2_SA)
adf.test(climate_data_ts$diff_LogCO2_SA[2:492])
```
(not sure how to interpret Kpss test, need to add things about the ACF plot) Since the p-value of the ADF test is less than 0.05, there is evidence to reject the null hypothesis in favor of the alternative hypothesis that the time series is stationary. 


```{r}
# choosing ARIMA model

result_dcmp_ARIMA_SNAIVE <- climate_data_ts %>%
  model(decomposition_model(
    STL(CO2 ~ season(window="periodic")),
    ARIMA(season_adjust ~ pdq(q=0) + PDQ(0,0,0),
          stepwise=FALSE, approximation=FALSE, trace=TRUE),
    SNAIVE(season_year)
    ))

report(result_dcmp_ARIMA_SNAIVE)
```